---
title: "Air-Quality-Trends"
output: html_notebook
---


```{r}
library(readr)
library(ggplot2)

data_url = "https://archive.ics.uci.edu/static/public/360/air+quality.zip"

download.file(data_url, destfile = "~/air_quality.zip", mode = "wb")
unzip("~/air_quality.zip")

data = read_delim("AirQualityUCI.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
data = data[, colSums(!is.na(data)) > 0]
head(data)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

## Variables
```{r}
str(data)
```


```{r}
names(data)
numeric_data <- data[, sapply(data, is.numeric)]


table <- data.frame(
  Variable = names(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Q1 = sapply(numeric_data, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) quantile(x, 0.5, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Q3 = sapply(numeric_data, function(x) quantile(x, 0.75, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE))
)

rownames(table) <- NULL

table
```
# Part 1: Exploratory Data Analysis

## Q1. Summary Statistics
```{r}
# Replace commas with dots and convert
data$`CO(GT)` <- as.numeric(gsub(",", ".", data$`CO(GT)`))
data$`C6H6(GT)` <- as.numeric(gsub(",", ".", data$`C6H6(GT)`))
data$T <- as.numeric(gsub(",", ".", data$T))
data$RH <- as.numeric(gsub(",", ".", data$RH))
data$AH <- as.numeric(gsub(",", ".", data$AH))
data[data == -200] <- NA
data$`CO(GT)` <- data$`CO(GT)` * 1000
# Rename columns to make them understandable
names(data) = c("date",
                "time",
                "co_ug_m3",
                "sensor_co",
                "nmhc_ug_m3",
                "benzene_ug_m3",
                "sensor_nmhc",
                "nox_ppb",
                "sensor_nox",
                "no2_ug_m3",
                "sensor_no2",
                "sensor_o3",
                "temperature_c",
                "relative_humidty",
                "absolute_humidity")

numeric_data <- data[, sapply(data, is.numeric)]

table <- data.frame(
  Variable = names(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Q1 = sapply(numeric_data, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) quantile(x, 0.5, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Q3 = sapply(numeric_data, function(x) quantile(x, 0.75, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE))
)

rownames(table) <- NULL

table
```

```{r}
str(data)
```



```{r}
# percentage of null values per column
sapply(data, function(col) {mean(col == -200, na.rm = TRUE) * 100})
```


```{r}
# Get the numerical columns from the data
numerical_variables = names(data)[sapply(data, is.numeric)]
numeric_data = data[, numerical_variables]

numerical_variables
numeric_data
```


```{r}
# Function to plot the distribution of numerical columns, 
# this column accepts a dataframe, the column name (name) and 
# a clean parameter to plot the distribution while replacing -200 based on the method that is chosen (True, False, Mean, Median, Mode, Random sample from sample space)

plot_numerical_distrbuition = function(df, name, clean = FALSE) {
  x = df[[name]]
  missing = which(x == -200)
  # For the clean value we can add the imputation method we are looking for
  if (clean == TRUE){ x = x[x != -200] }
  if (clean == "Mean") { x[missing] = mean(x[x != -200], na.rm = TRUE)}
  if (clean == "Median") { x[missing] = median(x[x != -200], na.rm = TRUE)}
  if (clean == "Mode") { 
    mode_values = names(sort(table(x[x != -200]), decreasing = TRUE))[1]
    x[missing] = as.numeric(mode_values)}
  if (clean == "Random") {
    set.seed(123)
    x[missing] = sample(x[x != -200], length(missing), replace = TRUE)}
  tmp = data.frame(value = x)
  
  ggplot(tmp, aes(x = value)) +
    geom_histogram(aes(y = ..density..),
                   bins = 30,
                   fill = "steelblue",
                   alpha = 0.7) +
    labs(title = paste("Distribution of", name),
         x = name,
         y = "Density") + 
    theme_minimal()
}
```


```{r}
# Plotting distribution of raw data
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col))
}
```


```{r}
# Plotting distribution replacing disregarding -200 values
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = TRUE))
}
```
```{r}
# Plotting distribution replacing -200 values with mean
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mean"))
}
```


```{r}
# Plotting distribution replacing -200 values with median
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Median"))
}
```


```{r}
# Plotting distribution replacing -200 values with mode
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mode"))
}
```
```{r}
# Plotting distribution replacing -200 values with random sample
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Random"))
}
```
```{r}
library(corrplot)
numeric_data = data[, sapply(data, is.numeric)]
numeric_data[numeric_data == -200] = NA
numeric_data = numeric_data[complete.cases(numeric_data),]

corr_matrix = cor(numeric_data)
corrplot(corr_matrix, method = "color", tl.cex = 0.8)

```


```{r}
pollutant_variables = c("co_ug_m3", "nmhc_ug_m3", "benzene_ug_m3", "nox_ppb", "no2_ug_m3")
data$datetime = as.POSIXct(paste(data$date, data$time), format = "%d/%m/%Y %H.%M.%S")
independent_variables = setdiff(names(data), c(pollutant_variables, "date", "time"))

## Median is defaulted as impute_data function
impute_data = function(df, method = "median") {
  df_copy <- df
  for (col in names(df_copy)) {
    if (is.numeric(df_copy[[col]])) {
      x <- df_copy[[col]]
      missing <- which(is.na(x))
      if (length(missing) > 0) {
        
        if (method == "mean") {
          x[missing] <- mean(x, na.rm = TRUE)
        }
        
        if (method == "median") {
          x[missing] <- median(x, na.rm = TRUE)
        }
        
        if (method == "mode") {
          mode_val <- as.numeric(names(sort(table(x), decreasing = TRUE)[1]))
          x[missing] <- mode_val
        }
        
        if (method == "random") {
          set.seed(123)
          x[missing] <- sample(x[!is.na(x)], length(missing), replace = TRUE)
        }
        
        df_copy[[col]] <- x
      }
    }
  }
  
  return(df_copy)
}
  

model_trends_function_impute = function(df, pollutant, variables, method = "median") {
  df <- impute_data(df, method = method)
  model_formula = as.formula(paste(pollutant, "~", paste(variables, collapse = "+")))
  cat("\nPollutant:", pollutant, "\n")
  cat("Impute Method:", method, "\n")
  model_trend = lm(model_formula, data = df)
  summary(model_trend)
}
```

```{r}
methods <- c("mean", "median", "mode", "random")

for (m in methods){
  for (col in pollutant_variables){
    cat("METHOD:", toupper(m), "\n")
    print(model_trends_function_impute(data, col, independent_variables, method = m))
  }
}
```



```{r}
data_impute_median <- impute_data(data, "median")
data_impute_mean <- impute_data(data, "mean")
data_impute_mode <- impute_data(data, "mode")
data_impute_random <- impute_data(data, "random")


ggplot(na.omit(data_impute_median), aes(x = datetime, y = co_ug_m3)) +
  geom_line() +
  theme_minimal() +
  labs(title = "CO Over Time (Impute median)")

ggplot(na.omit(data_impute_mean), aes(x = datetime, y = co_ug_m3)) +
  geom_line() +
  theme_minimal() +
  labs(title = "CO Over Time (Impute mean)")

ggplot(na.omit(data_impute_mode), aes(x = datetime, y = co_ug_m3)) +
  geom_line() +
  theme_minimal() +
  labs(title = "CO Over Time (Impute mode)")

ggplot(na.omit(data_impute_random), aes(x = datetime, y = co_ug_m3)) +
  geom_line() +
  theme_minimal() +
  labs(title = "CO Over Time (Impute random)")
```



```{r}
library(dplyr)
data_zoom <- data_impute_median %>%
  filter(datetime >= as.POSIXct("2004-03-01"),
         datetime <= as.POSIXct("2004-03-31"))

ggplot(na.omit(data_zoom), aes(datetime, co_ug_m3)) +
  geom_line() +
  theme_minimal()

data_zoom <- data_impute_median %>%
  filter(datetime >= as.POSIXct("2004-04-01"),
         datetime <= as.POSIXct("2004-04-30"))

ggplot(na.omit(data_zoom), aes(datetime, co_ug_m3)) +
  geom_line() +
  theme_minimal()

data_zoom <- data_impute_median %>%
  filter(datetime >= as.POSIXct("2004-05-01"),
         datetime <= as.POSIXct("2004-05-30"))

ggplot(na.omit(data_zoom), aes(datetime, co_ug_m3)) +
  geom_line() +
  theme_minimal()

data_zoom <- data_impute_median %>%
  filter(datetime >= as.POSIXct("2004-06-01"),
         datetime <= as.POSIXct("2004-06-30"))

ggplot(na.omit(data_zoom), aes(datetime, co_ug_m3)) +
  geom_line() +
  theme_minimal()
```


```{r}
library(dplyr)
data_impute <- data_impute %>%
  mutate(
    airstandard_co = 10,
    airstandard_nmhc = 300,
    airstandard_c6h6 = 5,
    airstandard_nox = 100,
    airstandard_no2 = 40
  )

data_impute <- data_impute %>%
  mutate(
    co_bad   = co_ug_m3 > airstandard_co,
    nmhc_bad = nmhc_ug_m3 > airstandard_nmhc,
    c6h6_bad = benzene_ug_m3 > airstandard_c6h6,
    nox_bad  = nox_ppb > airstandard_nox,
    no2_bad  = no2_ug_m3 > airstandard_no2
  )

data_impute <- data_impute %>%
  mutate(
    pollution_count = co_bad + nmhc_bad + c6h6_bad + nox_bad + no2_bad
  )

data_impute <- data_impute %>%
  mutate(
    air_quality = case_when(
      pollution_count == 0 ~ "Healthy",
      pollution_count == 1 ~ "Minor Pollution",
      pollution_count <= 3 ~ "Polluted",
      TRUE ~ "Major Pollution"
    )
  )

```

```{r}
ggplot(data_impute, aes(x = air_quality)) +
  geom_bar(fill = "steelblue") +
  theme_minimal()
```
```{r}
ggplot(data_impute, aes(datetime, fill = air_quality)) +
  geom_histogram(binwidth = 86400) +  # daily bins
  theme_minimal()
```


```{r}
sum(data_impute$exceed_8hr, na.rm = TRUE)
```