---
title: "Air-Quality-Trends"
output: html_notebook
---


```{r}
library(readr)
library(ggplot2)

data_url = "https://archive.ics.uci.edu/static/public/360/air+quality.zip"

download.file(data_url, destfile = "~/air_quality.zip", mode = "wb")
unzip("~/air_quality.zip")

data = read_delim("AirQualityUCI.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
data = data[, colSums(!is.na(data)) > 0]
head(data)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

## Variables
```{r}
str(data)
```


```{r}
names(data)
```
# Part 1: Exploratory Data Analysis

## Q1. Summary Statistics
```{r}
# Replace commas with dots and convert
data$`CO(GT)` <- as.numeric(gsub(",", ".", data$`CO(GT)`))
data$`C6H6(GT)` <- as.numeric(gsub(",", ".", data$`C6H6(GT)`))
data$T <- as.numeric(gsub(",", ".", data$T))
data$RH <- as.numeric(gsub(",", ".", data$RH))
data$AH <- as.numeric(gsub(",", ".", data$AH))

# Rename columns to make them understandable
names(data) = c("date",
                "time",
                "co_mg_m3",
                "sensor_co",
                "nmhc_ug_m3",
                "benzene_ug_m3",
                "sensor_nmhc",
                "nox_ppb",
                "sensor_nox",
                "no2_ug_m3",
                "sensor_no2",
                "sensor_o3",
                "temperature_c",
                "relative_humidty",
                "absolute_humidity")

summary(data)
```
```{r}
str(data)
```



```{r}
# percentage of null values per column
sapply(data, function(col) {mean(col == -200, na.rm = TRUE) * 100})
```


```{r}
# Get the numerical columns from the data
numerical_variables = names(data)[sapply(data, is.numeric)]
numerical_variables
```


```{r}
# Function to plot the distribution of numerical columns, 
# this column accepts a dataframe, the column name (name) and 
# a clean parameter to plot the distribution while replacing -200 based on the method that is chosen (True, False, Mean, Median, Mode, Random sample from sample space)

plot_numerical_distrbuition = function(df, name, clean = FALSE) {
  x = df[[name]]
  missing = which(x == -200)
  # For the clean value we can add the imputation method we are looking for
  if (clean == TRUE){ x = x[x != -200] }
  if (clean == "Mean") { x[missing] = mean(x[x != -200], na.rm = TRUE)}
  if (clean == "Median") { x[missing] = median(x[x != -200], na.rm = TRUE)}
  if (clean == "Mode") { 
    mode_values = names(sort(table(x[x != -200]), decreasing = TRUE))[1]
    x[missing] = as.numeric(mode_values)}
  if (clean == "Random") {
    set.seed(123)
    x[missing] = sample(x[x != -200], length(missing), replace = TRUE)}
  tmp = data.frame(value = x)
  
  ggplot(tmp, aes(x = value)) +
    geom_histogram(aes(y = ..density..),
                   bins = 30,
                   fill = "steelblue",
                   alpha = 0.7) +
    labs(title = paste("Distribution of", name),
         x = name,
         y = "Density") + 
    theme_minimal()
}
```


```{r}
# Plotting distribution of raw data
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col))
}
```


```{r}
# Plotting distribution replacing disregarding -200 values
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = TRUE))
}
```
```{r}
# Plotting distribution replacing -200 values with mean
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mean"))
}
```


```{r}
# Plotting distribution replacing -200 values with median
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Median"))
}
```


```{r}
# Plotting distribution replacing -200 values with mode
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mode"))
}
```
```{r}
# Plotting distribution replacing -200 values with random sample
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Random"))
}
```


```{r}


```


```{r}


```


```{r}

```


```{r}
```


```{r}
```


```{r}
```

