---
title: "Air-Quality-Trends"
author: "Hussein Ghosin, Kazuma Bowring, Derrick Li"
date: "2026-04-14"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE, comment=""}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE)
#Warning - false, message=false removes warning for package being installed
```


# Download and read in dataset
```{r, comment=""}
library(readr)
library(ggplot2)

data_url = "https://archive.ics.uci.edu/static/public/360/air+quality.zip"

download.file(data_url, destfile = "~/air_quality.zip", mode = "wb")
unzip("~/air_quality.zip")

data = read_delim("AirQualityUCI.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
data = data[, colSums(!is.na(data)) > 0]
head(data)
```

# Variables
```{r, comment=""}
str(data)
```

```{r}
dim(data)
```

# Summary of statistics prior to cleaning
```{r, comment=""}
names(data)
numeric_data <- data[, sapply(data, is.numeric)]

table <- data.frame(
  Variable = names(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Q1 = sapply(numeric_data, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) quantile(x, 0.5, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Q3 = sapply(numeric_data, function(x) quantile(x, 0.75, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE))
)
rownames(table) <- NULL

table
```

## Part 1: Exploratory Data Analysis

# Q1. Summary Statistics of Clean Data
```{r,comment=""}
# Replace commas with dots and convert
data$`CO(GT)` <- as.numeric(gsub(",", ".", data$`CO(GT)`))
data$`C6H6(GT)` <- as.numeric(gsub(",", ".", data$`C6H6(GT)`))
data$T <- as.numeric(gsub(",", ".", data$T))
data$RH <- as.numeric(gsub(",", ".", data$RH))
data$AH <- as.numeric(gsub(",", ".", data$AH))
data[data == -200] <- NA
data$`CO(GT)` <- data$`CO(GT)` * 1000

# Rename columns to make them understandable
names(data) = c("date",
                "time",
                "co_ug_m3",
                "sensor_co",
                "nmhc_ug_m3",
                "benzene_ug_m3",
                "sensor_nmhc",
                "nox_ppb",
                "sensor_nox",
                "no2_ug_m3",
                "sensor_no2",
                "sensor_o3",
                "temperature_c",
                "relative_humidty",
                "absolute_humidity")
data$nmhc_ug_m3 <- NULL
numeric_data <- data[, sapply(data, is.numeric)]
```

```{r, comment=""}
# Find null count for columns
null_table = data.frame(column = names(data), null_count = colSums(is.na(data)))
null_table
```


```{r, comment=""}
# Date and Time that are NA
null_col <- setdiff(names(data), c("date", "time"))

# Remove rows where both date and time are NA
data <- data[!apply(is.na(data[null_col]), 1, all), ]

# Sensors are NA
null_sens <- c("sensor_co", "sensor_nmhc", "sensor_nox", "sensor_no2", "sensor_o3")
 
# Remove rows where all sensors are NA
data <- data[!apply(is.na(data[null_sens]), 1, all), ]

# Remaining NA values will be imputed
null_count <- sapply(data, function(x) sum(is.na(x)))
null_count
```


```{r, comment=""}
clean_data <- data

clean_data[] <- lapply(clean_data, function(col){
  if (is.numeric(col)) {
    col[is.na(col)] <- median(col, na.rm = TRUE)
  }
  col
})
numerical_variables = names(clean_data)[sapply(clean_data, is.numeric)]
```

```{r, comment=""}
null_count <- sapply(clean_data, function(x) sum(is.na(x)))
null_count
```

```{r, comment=""}
clean_numeric_data <- clean_data[, sapply(clean_data, is.numeric)]

clean_table <- data.frame(
  Variable = names(clean_numeric_data),
  Min = sapply(clean_numeric_data, function(x) min(x, na.rm = TRUE)),
  Q1 = sapply(clean_numeric_data, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Median = sapply(clean_numeric_data, function(x) quantile(x, 0.5, na.rm = TRUE)),
  Mean = sapply(clean_numeric_data, function(x) mean(x, na.rm = TRUE)),
  Q3 = sapply(clean_numeric_data, function(x) quantile(x, 0.75, na.rm = TRUE)),
  Max = sapply(clean_numeric_data, function(x) max(x, na.rm = TRUE))
)
rownames(clean_table) <- NULL

clean_table
```


```{r, comment=""}
# Function to plot the distribution of numerical columns, 
# this column accepts a dataframe, the column name (name) and 
# a clean parameter to plot the distribution while replacing -200 based on the method that is chosen (True, False, Mean, Median, Mode, Random sample from sample space)

plot_numerical_distrbuition = function(df, name) {
  x = df[[name]]
  tmp = data.frame(value = x)
  
  ggplot(tmp, aes(x = value)) +
    geom_histogram(bins = 30,
                   fill = "steelblue",
                   alpha = 0.7) +
    labs(title = paste("Distribution of", name),
         x = name,
         y = "Density") + 
    theme_minimal()
}
```


# Q2. Effects of imputation on plots
```{r, echo=FALSE, comment=""}
# Plotting distribution replacing disregarding null values
for (col in numerical_variables){
  print(plot_numerical_distrbuition(clean_data, col))
}
```


```{r, echo=FALSE, comment=""}
library(corrplot)
numeric_data = clean_data[, sapply(data, is.numeric)]
numeric_data = numeric_data[complete.cases(numeric_data),]

corr_matrix = cor(numeric_data)
corrplot(corr_matrix, method = "color", tl.cex = 0.8)

```

```{r, comment=""}
pollutant_variables = c("co_ug_m3", "benzene_ug_m3", "nox_ppb", "no2_ug_m3")
clean_data$datetime = as.POSIXct(paste(data$date, data$time), format = "%d/%m/%Y %H.%M.%S")
independent_variables = setdiff(names(clean_data), c(pollutant_variables, "date", "time"))


model_trends_function = function(df, pollutant, variables) {
  
  model_formula = as.formula(paste(pollutant, "~", paste(variables, collapse = "+")))
  cat("\nPollutant:", pollutant, "\n")
  model_trend = lm(model_formula, data = df)
  summary(model_trend)
}
```

# Q3. Effects of Imputation of Summary
```{r, comment=""}
for (col in pollutant_variables){
  print(model_trends_function(clean_data, col, independent_variables))
}
```


## Part 2. Air Quality Index (AQI) Classification

```{r, comment=""}
clean_data$co_ppm = clean_data$co_ug_m3 * 24.45 / (28.01*1000)
clean_data$no2_ppb = clean_data$no2_ug_m3 * 24.45 / 46.01
clean_data$benzene_ppb = clean_data$benzene_ug_m3 * 24.45 / 78.11
```


```{r, comment=""}
breakpoints = data.frame(pollutant = c("co", "co", "co", "co","co", "co", "no2", "no2", "no2", "no2", "no2", "no2", "benzene", "benzene", "benzene", "benzene","benzene", "benzene", "nox", "nox","nox", "nox","nox", "nox"),
                         bp_low = c(0.0, 4.4, 9.4, 12.4, 15.4, 30.4, 0, 54, 101, 361, 650, 1250, 0, 5, 10, 20, 30, 50, 0, 53, 100, 360, 649, 1249),
                         bp_high = c(4.4, 9.4, 12.4, 15.4, 30.4, 40.4, 54, 101, 361, 650, 1250, 2049, 5, 10, 20, 30, 50, 80, 53, 100, 360, 649, 1249, 2049),
                         i_low = c(0, 51, 101, 151, 201, 301, 0, 51, 101, 151, 201, 301, 0, 51, 101, 151, 201, 301, 0, 51, 101, 151, 201, 301),
                         i_high = c(50, 100, 150, 200, 300, 500, 50, 100, 150, 200, 300, 500, 50, 100, 150, 200, 300, 500, 50, 100, 150, 200, 300, 500))
breakpoints
```


```{r, comment=""}
aqi_index = function(Cp, pollutant, breakpoints) {
  bp = breakpoints[breakpoints$pollutant == pollutant & Cp >= breakpoints$bp_low & Cp <= breakpoints$bp_high, ]
  
  if (nrow(bp) ==0 || is.na(Cp)) return(NA_real_)
  
  bp = bp[1, ]
  
  I = (((bp$i_high - bp$i_low) / (bp$bp_high - bp$bp_low)) * (Cp - bp$bp_low)) + bp$i_low
  return(I)
}
```


```{r, comment=""}
plot_health_over_time <- function(df, health_col = "aqi_health",
                                  time_unit = "day") {
  library(ggplot2)
  library(dplyr)
  library(scales)

  if (!health_col %in% names(df)) stop(paste("Column", health_col, "not found."))
  if (!"datetime" %in% names(df)) stop("Column 'datetime' not found.")

  plot_df <- df %>%
    select(datetime, health = all_of(health_col)) %>%
    filter(!is.na(health), !is.na(datetime))  # removes NA rows

  if (time_unit == "day") {
    plot_df <- plot_df %>% mutate(period = as.Date(datetime))
  } else if (time_unit == "week") {
    plot_df <- plot_df %>% mutate(period = as.Date(cut(as.Date(datetime), breaks = "week")))
  } else if (time_unit == "month") {
    plot_df <- plot_df %>% mutate(period = as.Date(format(datetime, "%Y-%m-01")))
  } else {
    stop("time_unit must be one of: 'day', 'week', 'month'")
  }

  summary_df <- plot_df %>%
    count(period, health) %>%
    group_by(period) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()

  ggplot(summary_df, aes(x = period, y = proportion, fill = health)) +
    geom_area(alpha = 0.8, position = "stack") +
    scale_fill_manual(values = c("healthy" = "steelblue", "not_healthy" = "tomato")) +  # fixed label
    scale_y_continuous(labels = percent_format()) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(
      title = paste("Health Status Over Time (by", time_unit, ")"),
      x     = "Date",
      y     = "Proportion",
      fill  = "Health Status"
    ) +
    theme_minimal() +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
}
```


```{r, comment=""}
plot_pollutants_by_health <- function(df, pollutants, health_col) {
  library(ggplot2)
  library(tidyr)
  
  long_df <- df %>%
    select(all_of(pollutants), all_of(health_col)) %>%
    pivot_longer(cols = all_of(pollutants),
                 names_to = "pollutant",
                 values_to = "value")
  
  ggplot(long_df, aes(x = value, fill = .data[[health_col]])) +
    geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
    facet_wrap(~ pollutant, scales = "free") +
    scale_fill_manual(values = c("healthy" = "forestgreen",
                                 "not_healthy" = "firebrick")) +
    labs(
      title = paste("Pollutant Distributions by", health_col),
      x = "Concentration",
      y = "Count",
      fill = "Health Status"
    ) +
    theme_minimal()
}

```




```{r, comment=""}
clean_data$i_co = sapply(clean_data$co_ppm, aqi_index, pollutant = "co", breakpoints = breakpoints)
clean_data$i_no2 = sapply(clean_data$no2_ppb, aqi_index, pollutant = "no2", breakpoints = breakpoints)
clean_data$i_benzene = sapply(clean_data$benzene_ppb, aqi_index, pollutant = "benzene", breakpoints = breakpoints)
clean_data$i_nox = sapply(clean_data$nox_ppb, aqi_index, pollutant = "nox", breakpoints = breakpoints)

clean_data$aqi = pmax(clean_data$i_co, clean_data$i_no2, clean_data$i_benzene, clean_data$i_nox, na.rm = TRUE)
clean_data$aqi_health = ifelse(clean_data$aqi <= 120, "healthy", "not_healthy")
clean_data
table(clean_data$aqi_health)

plot_health_over_time(clean_data, health_col = "aqi_health", time_unit = "month")

pollutants = c("co_ppm", "benzene_ppb", "nox_ppb", "no2_ppb")
plot_pollutants_by_health(clean_data, pollutants, "aqi_health")
```

# Q4. Model
```{r, echo=FALSE, comment=""}
#predicting AQI from only pollutants
set.seed(123)  # For reproducibility
aqi_index_trainingRowIndex <- sample(1:nrow(clean_data), 0.8 * nrow(data))  
aqi_index_train_data <- clean_data[aqi_index_trainingRowIndex, ]
aqi_index_test_data  <- clean_data[-aqi_index_trainingRowIndex, ]

model1_aqi_index <- lm(aqi ~ co_ppm+benzene_ppb+nox_ppb+no2_ppb, data = aqi_index_train_data)
#summary(model1_aqi_index)
predictions <- predict(model1_aqi_index, aqi_index_test_data)
#predictions

#Calculate CorrelationAccuracy 0 means bad or none, 1 is strong/good
actuals_preds <- data.frame(cbind(actual = (aqi_index_test_data$co_ppm+aqi_index_test_data$benzene_ppb+aqi_index_test_data$nox_ppb+aqi_index_test_data$no2_ppb), predicted = predictions))
correlation_accuracy <- cor(actuals_preds)
correlation_accuracy
#Calculate RMSE Root mean squared results vary base on variance of data range RMSE of 10 for 0-30 is bad, but good for 400-5000, being largely over the maximum value we get is ba
rmse <- sqrt(mean((actuals_preds$actual - actuals_preds$predicted)^2))
rmse
```


```{r, echo=FALSE, comment=""}
#predicting AQI from pollutants and interaction variables
set.seed(123)  # For reproducibility
aqi_index_trainingRowIndex_humidityntemp <- sample(1:nrow(clean_data), 0.8 * nrow(clean_data))  
aqi_index_train_data_humidityntemp <- clean_data[aqi_index_trainingRowIndex_humidityntemp, ]
aqi_index_test_data_humidityntemp  <- clean_data[-aqi_index_trainingRowIndex_humidityntemp, ]

model1_aqi_index_humidityntemp <- lm(aqi ~ co_ppm+benzene_ppb+nox_ppb+no2_ppb, data = aqi_index_train_data_humidityntemp)
#summary(model1_aqi_index)
predictions_humidityntemp <- predict(model1_aqi_index_humidityntemp, aqi_index_test_data_humidityntemp)
#predictions

#Calculate CorrelationAccuracy 0 means bad or none, 1 is strong/good
actuals_preds_humidityntemp <- data.frame(cbind(actual = (aqi_index_test_data_humidityntemp$co_ppm+
                                                       aqi_index_test_data_humidityntemp$benzene_ppb+
                                                       aqi_index_test_data_humidityntemp$nox_ppb+
                                                       aqi_index_test_data_humidityntemp$no2_ppb+
                                                       aqi_index_test_data_humidityntemp$temperature_c+
                                                       aqi_index_test_data_humidityntemp$relative_humidty+
                                                       aqi_index_test_data_humidityntemp$absolute_humidity)^2, predicted = predictions))
correlation_accuracy_humidityntemp <- cor(actuals_preds_humidityntemp)
correlation_accuracy_humidityntemp
#Calculate RMSE Root mean squared results vary base on variance of data range RMSE of 10 for 0-30 is bad, but good for 400-5000, being largely over the maximum value we get is ba
rmse <- sqrt(mean((actuals_preds_humidityntemp$actual - actuals_preds_humidityntemp$predicted)^2))
rmse

```


```{r, echo=FALSE, comment=""}
#DO note AIC's may be large simply because our sample is massive, so a large VIF doesnt mean its bad
#generalized linear model predicting aqi_health in response to pollutants omitting n/a results and predictors
#str(data)
data_remove_outlier <-  clean_data
glm_fit_pollutants_aqi_healthy<- glm(as.factor(clean_data$aqi_health) ~ co_ug_m3+benzene_ppb+nox_ppb+no2_ug_m3, data = data_remove_outlier,family = binomial,na.action = na.omit)
summary(glm_fit_pollutants_aqi_healthy)
plot(glm_fit_pollutants_aqi_healthy)
aic_glm_fit_pollutants_aqi_healthy<- AIC(glm_fit_pollutants_aqi_healthy)
print(paste("the AIC value of this GLM is ", aic_glm_fit_pollutants_aqi_healthy))
```


```{r, echo=FALSE, comment=""}
#linear model predicting aqi based on aqi of pollutants
lm_fit_aqi_pollutants_aqi <- lm(aqi ~ co_ppm+no2_ppb+i_co+i_no2+i_benzene+i_nox, data = clean_data, na.action = na.omit)
summary(lm_fit_aqi_pollutants_aqi)
plot(lm_fit_aqi_pollutants_aqi)
aic_lm_aqi_pollutants_aqi <- AIC(lm_fit_aqi_pollutants_aqi)
print(paste("the AIC value of this LM (AQI prediction based on pollutant AQI) is ", aic_lm_aqi_pollutants_aqi))
```

```{r, echo=FALSE, comment=""}
#linear model predicting aqi based on pollutants omitting na values from linear modelling,
lm_fit_aqi_pollutants_aqi <- lm(aqi ~ co_ppm+benzene_ppb+nox_ppb+no2_ppb, data = clean_data, na.action = na.omit)
summary(lm_fit_aqi_pollutants_aqi)
plot(lm_fit_aqi_pollutants_aqi)
aic_lm_aqi_pollutants_aqi <- AIC(lm_fit_aqi_pollutants_aqi)
print(paste("the AIC value of this LM (AQI prediction based on pollutant ) is ", aic_lm_aqi_pollutants_aqi))
```

```{r, echo=FALSE, comment=""}
#for VIF
#install.packages("car")
library(car)

#linear model predicting aqi based on pollutants and temperature + both humidities ^2 to chart all interaction variables
lm_fit_aqi_vs_pollutants_temp_humid <- lm(aqi ~ (co_ppm+benzene_ppb+nox_ppb+no2_ppb+temperature_c+relative_humidty+absolute_humidity)^2, data = clean_data, na.action = na.omit)
summary(lm_fit_aqi_vs_pollutants_temp_humid)
plot(lm_fit_aqi_vs_pollutants_temp_humid)
aic_lm_aqi_pollutants_humidity_aq <- AIC(lm_fit_aqi_vs_pollutants_temp_humid)
print(paste("the AIC value of this LM (AQI prediction based on pollutant and temperature + humidities) is ", aic_lm_aqi_pollutants_humidity_aq))
vif_lm_fit_aqi_vs_pollutants_temp_humid <- vif(lm_fit_aqi_vs_pollutants_temp_humid,type=c("predictor"))
print(paste("the GVIF^(1/(2*Df)) value of this LM (AQI prediction based on pollutant and temperature + humidities) is 1"))
#VIF's return 1 for GVIF^(1/(2*Df)) = low multicollinearity
#Vif's return 1 as well for GVIF, this is our RAW VIF
```

```{r, echo=FALSE, comment=""}
#VIF library (car) in previous code block

# GENERALIZED linear model predicting aqi based on pollutants and temperature + both humidities ^2 to chart all interaction variables
glm_fit_aqi_vs_pollutants_temp_humid <- glm(as.factor(clean_data$aqi_health) ~ (co_ppm+benzene_ppb+nox_ppb+no2_ppb+temperature_c+relative_humidty+absolute_humidity)^2, data = clean_data, family = binomial,na.action = na.omit)
summary(glm_fit_aqi_vs_pollutants_temp_humid)
plot(glm_fit_aqi_vs_pollutants_temp_humid)
aic_glm_aqi_pollutants_aqi <- AIC(glm_fit_aqi_vs_pollutants_temp_humid)
#vif_glm_fit_aqi_vs_pollutants_temp_humid <- vif(glm_fit_aqi_vs_pollutants_temp_humid,type=c("predictor"))

```

```{r, echo=FALSE, comment=""}
#linear model predicting aqi based on pollutants omitting na values from linear modelling,
lm_fit_aqi_pollutants_aqi <- lm(aqi ~ co_ppm+benzene_ppb+nox_ppb+no2_ppb, data = clean_data, na.action = na.omit)
summary(lm_fit_aqi_pollutants_aqi)
plot(lm_fit_aqi_pollutants_aqi)
aic_lm_aqi_pollutants_aqi <- AIC(lm_fit_aqi_pollutants_aqi)
```

```{r, echo=FALSE, comment=""}
#for VIF
#install.packages("car")
library(car)

#linear model predicting aqi based on pollutants and temperature + both humidities ^2 to chart all interaction variables
lm_fit_aqi_vs_pollutants_temp_humid <- lm(aqi ~ (co_ppm+benzene_ppb+nox_ppb+no2_ppb+temperature_c+relative_humidty+absolute_humidity)^2, data = clean_data, na.action = na.omit)
summary(lm_fit_aqi_vs_pollutants_temp_humid)
plot(lm_fit_aqi_vs_pollutants_temp_humid)
aic_lm_aqi_pollutants_aqi <- AIC(lm_fit_aqi_vs_pollutants_temp_humid)
vif_lm_fit_aqi_vs_pollutants_temp_humid <- vif(lm_fit_aqi_vs_pollutants_temp_humid,type=c("predictor"))
#VIF's return 1 for GVIF^(1/(2*Df)) =(adjusted (normalized) GVIF) which is good as a GVIF^(1/(2*Df)) >2.2/3.2 means theres a problem >1.6 = concern
#Vif's return 1 as well for GVIF, this is our RAW VIF
```

```{r, echo=FALSE, comment=""}
#VIF library (car) in previous code block

# GENERALIZED linear model predicting aqi based on pollutants and temperature + both humidities ^2 to chart all interaction variables
glm_fit_aqi_vs_pollutants_temp_humid <- glm(as.factor(clean_data$aqi_health) ~ (co_ppm+benzene_ppb+nox_ppb+no2_ppb+temperature_c+relative_humidty+absolute_humidity)^2, data = clean_data, family = binomial,na.action = na.omit)
summary(glm_fit_aqi_vs_pollutants_temp_humid)
plot(glm_fit_aqi_vs_pollutants_temp_humid)
aic_glm_aqi_pollutants_aqi <- AIC(glm_fit_aqi_vs_pollutants_temp_humid)
vif_glm_fit_aqi_vs_pollutants_temp_humid <- vif(glm_fit_aqi_vs_pollutants_temp_humid,type=c("predictor"))
#VIF's return 1 for GVIF^(1/(2*Df)) =(adjusted (normalized) GVIF) which is good as a GVIF^(1/(2*Df)) >2.2/3.2 means theres a problem >1.6 = concern
#Vif's return 1 as well for GVIF, this is our RAW VIF
```

# Q5. Sample Hypothesis'
```{r}
#Splitting Data
set.seed(42)
sample_index <- sample(1:nrow(clean_data), 0.8 * nrow(clean_data))
train_data <- clean_data[sample_index, ]
test_data  <- clean_data[-sample_index, ]
```

# H_0 (null): Pollutant concentrations have no effect on the calculated AQI
# H_a (alternative): At least one pollutant concentration significantly predicts the AQI
```{r}
# Hypothesis 1: Pollutant variables Model
m1 <- lm(log(aqi) ~ co_ppm + benzene_ppb + nox_ppb + no2_ppb, 
         data = train_data)

pred1 <- predict(m1, newdata = test_data)

rmse1 <- sqrt(mean((test_data$aqi - pred1)^2))
rmse1

mae1 <- mean(abs(test_data$aqi - pred1))
mae1

sse1 <- sum((test_data$aqi - pred1)^2)
sst1 <- sum((test_data$aqi - mean(test_data$aqi))^2)
r2_test1 <- 1 - sse1/sst1
r2_test1

AIC(m1)
summary(m1)
par(mfrow = c(2, 2))
plot(m1)
```
```{r}
anova_table1 <- anova(m1)
print(anova_table1)
```

# H_0: Environmental factors have no relationship with log(AQI)
# H_a: Environmetal factors significantly influence log(AQI)
```{r}
# Hypothesis 2 Environmental variables 
m2 <- lm(log(aqi) ~ temperature_c + relative_humidty + absolute_humidity, 
         data = train_data)

pred2 <- exp(predict(m2, newdata = test_data))

rmse2 <- sqrt(mean((test_data$aqi - pred2)^2))
rmse2

mae2 <- mean(abs(test_data$aqi - pred2))
mae2

sse2 <- sum((test_data$aqi - pred2)^2)
sst2 <- sum((test_data$aqi - mean(test_data$aqi))^2)
r2_test2 <- 1 - sse2/sst2
r2_test2

AIC(m2)
summary(m2)
par(mfrow = c(2, 2))
plot(m2)
```
```{r}
anova_table2 <- anova(m2)
print(anova_table2)
```


# H_0: Combining environmental variables and pollutant concentrations does not improve the models predictions
# H_a: The combined model provides statistically more fit AQI predicting
```{r}
# Hypothesis 3: Combination of environmental and pollutant variables
m3 <- lm(log(aqi) ~ co_ppm + benzene_ppb + nox_ppb + no2_ppb +
               temperature_c + relative_humidty + absolute_humidity, 
         data = train_data)

pred3 <- exp(predict(m3, newdata = test_data))

rmse3 <- sqrt(mean((test_data$aqi - pred3)^2))
rmse3

mae3 <- mean(abs(test_data$aqi - pred3))
mae3

sse3 <- sum((test_data$aqi - pred1)^2)
sst3 <- sum((test_data$aqi - mean(test_data$aqi))^2)
r2_test3 <- 1 - sse3/sst3
r2_test3

AIC(m3)
summary(m3)
par(mfrow = c(2, 2))
plot(m3)
```


```{r}
anova_table <- anova(m3)
print(anova_table)
```

```{r}
library(knitr)

results <- data.frame(Model = c("M1", "M2", "M3"),
                      RMSE = c(rmse1, rmse2, rmse3),
                      MAE = c(mae1, mae2, mae3),
                      Test_R2 = c(r2_test1, r2_test2, r2_test3),
                      AIC = c(AIC(m1), AIC(m2), AIC(m3)))

kable(results, caption = "Metrics for All Models", digits = 4)
```

