---
title: "Air-Quality-Trends"
output: html_notebook
---


```{r}
library(readr)
library(ggplot2)

data_url = "https://archive.ics.uci.edu/static/public/360/air+quality.zip"

download.file(data_url, destfile = "~/air_quality.zip", mode = "wb")
unzip("~/air_quality.zip")

data = read_delim("AirQualityUCI.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
data = data[, colSums(!is.na(data)) > 0]
head(data)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

## Variables
```{r}
str(data)
```


```{r}
names(data)
```
# Part 1: Exploratory Data Analysis

## Q1. Summary Statistics
```{r}
# Replace commas with dots and convert
data$`CO(GT)` <- as.numeric(gsub(",", ".", data$`CO(GT)`))
data$`C6H6(GT)` <- as.numeric(gsub(",", ".", data$`C6H6(GT)`))
data$T <- as.numeric(gsub(",", ".", data$T))
data$RH <- as.numeric(gsub(",", ".", data$RH))
data$AH <- as.numeric(gsub(",", ".", data$AH))

# Rename columns to make them understandable
names(data) = c("date",
                "time",
                "co_mg_m3",
                "sensor_co",
                "nmhc_ug_m3",
                "benzene_ug_m3",
                "sensor_nmhc",
                "nox_ppb",
                "sensor_nox",
                "no2_ug_m3",
                "sensor_no2",
                "sensor_o3",
                "temperature_c",
                "relative_humidty",
                "absolute_humidity")

summary(data)
```
```{r}
str(data)
```



```{r}
# percentage of null values per column
sapply(data, function(col) {mean(col == -200, na.rm = TRUE) * 100})
```


```{r}
# Get the numerical columns from the data
numerical_variables = names(data)[sapply(data, is.numeric)]
numeric_data = data[, numerical_variables]

numerical_variables
numeric_data
```


```{r}
# Function to plot the distribution of numerical columns, 
# this column accepts a dataframe, the column name (name) and 
# a clean parameter to plot the distribution while replacing -200 based on the method that is chosen (True, False, Mean, Median, Mode, Random sample from sample space)

plot_numerical_distrbuition = function(df, name, clean = FALSE) {
  x = df[[name]]
  missing = which(x == -200)
  # For the clean value we can add the imputation method we are looking for
  if (clean == TRUE){ x = x[x != -200] }
  if (clean == "Mean") { x[missing] = mean(x[x != -200], na.rm = TRUE)}
  if (clean == "Median") { x[missing] = median(x[x != -200], na.rm = TRUE)}
  if (clean == "Mode") { 
    mode_values = names(sort(table(x[x != -200]), decreasing = TRUE))[1]
    x[missing] = as.numeric(mode_values)}
  if (clean == "Random") {
    set.seed(123)
    x[missing] = sample(x[x != -200], length(missing), replace = TRUE)}
  tmp = data.frame(value = x)
  
  ggplot(tmp, aes(x = value)) +
    geom_histogram(aes(y = ..density..),
                   bins = 30,
                   fill = "steelblue",
                   alpha = 0.7) +
    labs(title = paste("Distribution of", name),
         x = name,
         y = "Density") + 
    theme_minimal()
}
```


```{r}
# Plotting distribution of raw data
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col))
}
```


```{r}
# Plotting distribution replacing disregarding -200 values
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = TRUE))
}
```
```{r}
# Plotting distribution replacing -200 values with mean
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mean"))
}
```


```{r}
# Plotting distribution replacing -200 values with median
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Median"))
}
```


```{r}
# Plotting distribution replacing -200 values with mode
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Mode"))
}
```
```{r}
# Plotting distribution replacing -200 values with random sample
for (col in numerical_variables){
  print(plot_numerical_distrbuition(data, col, clean = "Random"))
}
```
```{r}
library(corrplot)
numeric_data = data[, sapply(data, is.numeric)]
numeric_data[numeric_data == -200] = NA
numeric_data = numeric_data[complete.cases(numeric_data),]

corr_matrix = cor(numeric_data)
corrplot(corr_matrix, method = "color", tl.cex = 0.8)

```
```{r}
pollutant_variables = c("co_mg_m3", "nmhc_ug_m3", "benzene_ug_m3", "nox_ppb", "no2_ug_m3")
data$datetime = as.POSIXct(paste(data$date, data$time), format = "%d/%m/%Y %H.%M.%S")
independent_variables = setdiff(names(data), c(pollutant_variables, "date", "time"))
```


```{r}


fill_missing_values = function(df, method = FALSE){
  if (method == FALSE) return(df)
  
  set.seed(123)
  
  for (col in names(df)){
    x = df[[col]]
    if (is.numeric(x)){
      missing = which(x == -200)
      
      if (length(missing) > 0) {
        # For the method value we can add the imputation method we are looking for
        if (method == "Mean") { df[[col]][missing] = mean(x[x != -200 || x != -174.5805070], na.rm = TRUE)}
        if (method == "Median") { df[[col]][missing] = median(x[x != -200], na.rm = TRUE)}
        if (method == "Mode") { 
          mode_values = names(sort(table(x[x != -200]), decreasing = TRUE))[1]
          df[[col]][missing] = as.numeric(mode_values)}
        if (method == "Random") {
          set.seed(123)
          df[[col]][missing] = sample(x[x != -200], length(missing), replace = TRUE)}
      }
    }
  }
return(df)
}
  
  

model_trends_function = function(df, pollutant, variables, clean = FALSE) {
  df = fill_missing_values(df, method = clean)
  if (any(df == -200, na.rm = TRUE) == TRUE & clean != FALSE) return(print("there are still missing values"))
  model_formula = as.formula(paste(pollutant, "~", paste(variables, collapse = "+")))
  print(pollutant)
  print(clean)
  model_trend = lm(model_formula,
                   data = df)
  summary(model_trend)}


for (col in pollutant_variables){
  print(model_trends_function(data, col, independent_variables))
}

```


```{r}

for (col in pollutant_variables){
  print(model_trends_function(data, col, independent_variables, clean = "Mean"))
}
```


```{r}
for (col in pollutant_variables){
  print(model_trends_function(data, col, independent_variables, clean = "Mode"))
}
```


```{r}
for (col in pollutant_variables){
  print(model_trends_function(data, col, independent_variables, clean = "Median"))
}

```


```{r}
for (col in pollutant_variables){
  print(model_trends_function(data, col, independent_variables, clean = "Random"))
}


```


```{r}
replace_missing_values = function(x) {
  valid_mean = mean(x[x != -200], na.rm = TRUE)
  x[x == -200] = valid_mean
  return(x)
}
for (poll in pollutant_variables) {
  data[[poll]] = replace_missing_values(data[[poll]])
}
data$co_ppm = data$co_mg_m3 * 24.45 / 28.01
data$no2_ppb = data$no2_ug_m3 * 24.45 / 46.01
```


```{r}
breakpoints = data.frame(pollutant = c("co", "co", "no2", "no2", "benzene", "benzene", "nox", "nox"),
                         bp_low = c(0.0, 4.5, 0, 54, 0, 1, 0, 54),
                         bp_high = c(4.4, 9.4, 53, 100, 1, 3,  53, 100),
                         i_low = c(0, 51, 0, 51, 0, 51, 0, 51),
                         i_high = c(50, 100, 50, 100, 50, 100, 50, 100))
breakpoints
```

```{r}

aqi_index = function(Cp, pollutant, breakpoints) {
  bp = breakpoints[breakpoints$pollutant == pollutant & Cp >= breakpoints$bp_low & Cp <= breakpoints$bp_high, ]
  
  if (nrow(bp) ==0 || is.na(Cp)) return(NA_real_)
  
  bp = bp[1, ]
  
  I = (((bp$i_high - bp$i_low) / (bp$bp_high - bp$bp_low)) * (Cp - bp$bp_low)) + bp$i_low
  return(I)
}
```




```{r}
plot_health_over_time <- function(df, health_col = "aqi_health",
                                  time_unit = "day") {
  library(ggplot2)
  library(dplyr)
  library(scales)

  if (!health_col %in% names(df)) stop(paste("Column", health_col, "not found."))
  if (!"datetime" %in% names(df)) stop("Column 'datetime' not found.")

  plot_df <- df %>%
    select(datetime, health = all_of(health_col)) %>%
    filter(!is.na(health), !is.na(datetime))  # removes NA rows

  if (time_unit == "day") {
    plot_df <- plot_df %>% mutate(period = as.Date(datetime))
  } else if (time_unit == "week") {
    plot_df <- plot_df %>% mutate(period = as.Date(cut(as.Date(datetime), breaks = "week")))
  } else if (time_unit == "month") {
    plot_df <- plot_df %>% mutate(period = as.Date(format(datetime, "%Y-%m-01")))
  } else {
    stop("time_unit must be one of: 'day', 'week', 'month'")
  }

  summary_df <- plot_df %>%
    count(period, health) %>%
    group_by(period) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()

  ggplot(summary_df, aes(x = period, y = proportion, fill = health)) +
    geom_area(alpha = 0.8, position = "stack") +
    scale_fill_manual(values = c("healthy" = "steelblue", "not_healthy" = "tomato")) +  # fixed label
    scale_y_continuous(labels = percent_format()) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(
      title = paste("Health Status Over Time (by", time_unit, ")"),
      x     = "Date",
      y     = "Proportion",
      fill  = "Health Status"
    ) +
    theme_minimal() +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
}
```


```{r}
unique(data$aqi_health)
unique(data$threshold_health)
```
```{r}
plot_pollutants_by_health <- function(df, pollutants, health_col) {
  library(ggplot2)
  library(tidyr)
  
  long_df <- df %>%
    select(all_of(pollutants), all_of(health_col)) %>%
    pivot_longer(cols = all_of(pollutants),
                 names_to = "pollutant",
                 values_to = "value")
  
  ggplot(long_df, aes(x = value, fill = .data[[health_col]])) +
    geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
    facet_wrap(~ pollutant, scales = "free") +
    scale_fill_manual(values = c("healthy" = "forestgreen",
                                 "not_healthy" = "firebrick")) +
    labs(
      title = paste("Pollutant Distributions by", health_col),
      x = "Concentration",
      y = "Count",
      fill = "Health Status"
    ) +
    theme_minimal()
}

```



```{r}
data$i_co = sapply(data$co_ppm, aqi_index, pollutant = "co", breakpoints = breakpoints)
data$i_no2 = sapply(data$no2_ppb, aqi_index, pollutant = "no2", breakpoints = breakpoints)
data$i_benzene = sapply(data$benzene_ug_m3, aqi_index, pollutant = "benzene", breakpoints = breakpoints)
data$i_nox = sapply(data$nox_ppb, aqi_index, pollutant = "nox", breakpoints = breakpoints)

data$aqi = pmax(data$i_co, data$i_no2, data$i_benzene, data$i_nox, na.rm = TRUE)
data$aqi_health = ifelse(data$aqi <= 70, "healthy", "not_healthy")

table(data$aqi_health)

plot_health_over_time(data, health_col = "aqi_health", time_unit = "month")

pollutants = c("co_ppm", "benzene_ug_m3", "nox_ppb", "no2_ppb")
plot_pollutants_by_health(data, pollutants, "aqi_health")
```


```{r}
data$healthy_co = ifelse(data$co_mg_m3 < 10, 1, 0)
data$healthy_benzene = ifelse(data$benzene_ug_m3 < 5, 1, 0)
data$healthy_nox = ifelse(data$nox_ppb < 53, 1, 0)
data$healthy_no2 = ifelse(data$no2_ug_m3 < 53, 1, 0)

data$health_score = data$healthy_benzene + data$healthy_co + data$healthy_no2 + data$healthy_nox
data$threshold_health = ifelse(data$health_score >= 2, "healthy", "not_healthy")

table(data$threshold_health)

plot_health_over_time(data, health_col = "threshold_health", time_unit = "month")

plot_pollutants_by_health(data, pollutants, "threshold_health")
```

